Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-09-27 astronum_2018/astronum-xrb.tex
minor edit
fix acks
clean up some descriptions

One-dimensional simulations, e.g. \cite{woosley-xrb,fisker:2008}, assume
spherical symmetry and model the burst through the vertical column
depth of fuel.  These simulations can capture the luminosities and
burst recurrence times well.  They can utilize very large nuclear
reaction networks and have been used to understand the nucleosynthesis
and rp-process~\cite{shatz_rp}, as well as to understand how rate uncertainties can
affect the outcomes of the bursts.  Many successive bursts can be
modeled and these simulations can show us the evolution of the ash
layer with time.  The approximation of 1D however means that we cannot
learn about lateral variations across the neutron star, including
flame spreading.

@@

A longstanding goal is to perform simulations where we resolve the
flame structure, to allow us to accurately capture the
nucleosynthesis, and watch it move across the neutron star surface.
The results of these simulations will enable strict comparison to
observations made with the new generation of X-ray telescopes such as
\textit{eXTP} and \textit{STROBE--X}
\cite{art-2017-wilhod-etal,art-2016-zhang-etal} which will offer
high collecting area and time resolution.
We show some in-progress calculations that are a step toward this, discuss
their remaining approximations, and also discuss what new algorithmic
techniques might be needed to realize this goal with the advent of
exascale computing.

@@

We setup a problem by creating a pair of 1D hydrostatic XRB models
representing the post- and pre-flame states and apply these laterally
across the domain to setup a hot region at the left end of our domain
in equilibrium with the cool model.  Thermal diffusion and nuclear
reactions quickly setup a laterally propagating flame that begins to
move through our domain.  We resolve the diffusion scale, using a
resolution of 10~cm, and use adaptive mesh refinement (two jumps of
$4\times$) to keep a buffer of low density material at low resolution
above the surface to allow for expansion.  To make the lengthscales
tractable, we rotate the neutron star at 2000~Hz---much faster than
expected.  This helps confine the spreading region to a more
reasonable horizontal region.  Nevertheless, our simulation uses 12288
zones laterally, on the finest grid.  We use the stellar
conducitivities from \cite{Timmes00} and a 13-isotopic alpha-chain
reaction network.  To make the timescales easier, we boost the flame
speed by a factor of 10, by boosting both the conductivity and energy
generation rate.  This is an approximation we hope to relax in the
near future.

Figure~\ref{fig:xrb} shows the flame structure for an initial 2D run
at 0.007~s.  The burning layer is only about 10~m high.  Underneath is
\isot{Ni}{56}, representative of the underlying neutron star, and
above it is very a low density ($10^{-4}~\gcc$) buffer.  Only a
portion of the vertical extent is show.  The top panel shows the
temperature.  The initial perturbation reached to $2\times
10^4~\mathrm{cm}$, so at this point, the extent of the burnt region
has nearly doubled.  The flame structure is quite wide, since there
are multiple burning stages beyond just helium to carbon, and this
shows up in the composition.  This is seen more clearly in the lower
panel which shows the mean molecular weight of the nuclei ($\bar{A}$).
We see the further behind the head of the flame we are, the heavier
the ash nuclei.  The second panel shows the energy generation rate.
It is strongest just above the neutron star surface, behind the flame.
We also see that the head of the flame is ``detached'' from the base
of the atmosphere, with the strongest burning at the head of the flame
slightly higher up in the atmosphere.  This can also be more clearly
seen in the last panel that shows the ash structure.  This appears
reminiscent of the flame structure seen in the earlier simulations of
\cite{cavecchi:2012} \MarginPar{Yuri, would you agree?}  The third
panel shows the velocity through the plane of the simulation---this is
induced by the Coriolis force as the burning front spreads laterally.
A tight hurricane structure has setup as a result of the flame
spreading.  This is the geostrophic balance discussed in
\cite{SPIT_ETAL02}.  Finally, as already mentioned, the last panel
shows the ash composition.  These calculations are ongoing and will be
the subject of a detailed study in the near future.

\begin{figure}
\centering
\includegraphics[width=0.95\linewidth]{xrb}
\caption{\label{fig:xrb} Helium flame spreading across the surface of
  a neutron star.  Shown are temperature (top), the velocity out of the
  simulation plane (middle), and nuclear energy generation rate
  (bottom).
  The out of plane velocity contours indicate where the Coriolis force
  has created a hurricane-like wind structure which confines the flame,
  leading to an inclined front. It is at such an interface that the
  temperature and burning rate are highest.
  }
\end{figure}

@@

\ack The work at Stony Brook was supported by DOE/Office of Nuclear
Physics grant DE-FG02-87ER40317 and contract 7418390 with Lawrence
Berkeley National Laboratory as part of the Exascale Compute Project
ExaStar collaboration.  The work at LBNL was supported by the DOE
Office of Advanced Scientific Computing Research under Contract No,
DE-AC02-05CH11231. An award of computer time was provided by the
Innovative and Novel Computational Impact on Theory and Experiment
(INCITE) program. YC is supported by the European Union Horizon 2020
research and innovation programme under the Marie Sklodowska-Curie
Global Fellowship grant agreement No 703916. This research used
resources of the Oak Ridge Leadership Computing Facility at the Oak
Ridge National Laboratory, which is supported by the Office of Science
of the U.S. Department of Energy under Contract
No.\ DE-AC05-00OR22725.  This research used resources of the National
Energy Research Scientific Computing Center, which is supported by the
Office of Science of the U.S. Department of Energy under Contract
No.\ DE-AC02-05CH11231.  Visualizations were done using yt~\cite{yt}.
This research has made use of NASA's Astrophysics Data System
Bibliographic Services.

>>>>

One-dimensional simulations, e.g. \cite{woosley-xrb,fisker:2008}, assume
spherical symmetry and model the burst through the vertical column
depth of fuel.  These simulations can capture the luminosities and
burst recurrence times well.  They can utilize very large nuclear
reaction networks and have been used to understand the nucleosynthesis
and rp-process~\cite{schatz:rp1999,rpprocess,schatz_rp}, as well as to understand how rate uncertainties can
affect the outcomes of the bursts.  Many successive bursts can be
modeled and these simulations can show us the evolution of the ash
layer with time.  The approximation of 1D however means that we cannot
learn about lateral variations across the neutron star, including
flame spreading.

@@

A longstanding goal is to perform simulations where we resolve the
flame structure, to allow us to accurately capture the
nucleosynthesis, and watch it move across the neutron star surface.
The results of these simulations will enable strict comparison to
observations made with the new generation of X-ray telescopes such as
\textit{eXTP}\/ and \textit{STROBE-X}\/
\cite{art-2017-wilhod-etal,art-2016-zhang-etal} which will offer
high collecting area and time resolution.
We show some in-progress calculations that are a step toward this, discuss
their remaining approximations, and also discuss what new algorithmic
techniques might be needed to realize this goal with the advent of
exascale computing.

@@

We setup a problem by creating a pair of 1D hydrostatic XRB models (hot and cool)
representing the post- and pre-flame states and apply these laterally
across the domain to setup a hot region at the left end of our domain
in equilibrium with the cool model.  Thermal diffusion and nuclear
reactions quickly setup a laterally propagating flame that begins to
move through our domain.  We resolve the diffusion scale, using a
resolution of 10~cm, and use adaptive mesh refinement (two jumps of
$4\times$) to keep a buffer of low density material at low resolution
above the surface to allow for expansion.  To make the lengthscales
tractable, we rotate the neutron star at 2000~Hz---much faster than
expected.  This helps confine the spreading region to a more
reasonable horizontal region.  Nevertheless, our simulation uses 12288
zones laterally, on the finest grid.  We use the stellar
conducitivities from \cite{Timmes00} and a 13-isotopic alpha-chain
reaction network.  To make the timescales easier, we boost the flame
speed by a factor of 10, by boosting both the conductivity and energy
generation rate.  This is an approximation we hope to relax in the
near future.

Figure~\ref{fig:xrb} shows the flame structure for an initial 2D run
at 0.007~s.  The burning layer is only about 10~m high.  Underneath is
\isot{Ni}{56}, representative of the underlying neutron star, and
above it is very a low density ($10^{-4}~\gcc$) buffer.  Only a
portion of the vertical extent is show.  The top panel shows the
temperature structure.  The initial perturbation reached to $2\times
10^4~\mathrm{cm}$, so at this point, the extent of the burnt region
has nearly doubled.  The flame structure is quite wide, since there
are multiple burning stages beyond just helium to carbon, and this
shows up in the composition.  This is seen more clearly in the next 
panel which shows the mean molecular weight of the nuclei ($\bar{A}$).
We see the further behind the head of the flame we are, the heavier
the ash nuclei.  We also see where that the head of the flame
appears lifted off of the base of the burning layer---unburned
material underlies the spreading flame.  This appears
reminiscent of the flame structure seen in the earlier simulations of
\cite{cavecchi:2012}. \MarginPar{Yuri, would you agree?} The third panel shows the energy generation rate.
It is strongest just above the neutron star surface, behind the flame.
We again see that the head of the flame is detached from the base
of the atmosphere, with the strongest burning at the head of the flame
slightly higher up in the atmosphere.   Finally, the bottom
panel shows the velocity through the plane of the simulation---this is
induced by the Coriolis force as the burning front spreads laterally.
A tight hurricane structure has setup as a result of the flame
spreading.  This is the geostrophic balance discussed in
\cite{SPIT_ETAL02}.    These calculations are ongoing and will be
the subject of a detailed study in the near future.

\begin{figure}
\centering
\includegraphics[width=0.95\linewidth]{xrb}
\caption{\label{fig:xrb} Helium flame spreading across the surface of
  a neutron star.  Shown are temperature (top), the mean molecular
  weight of the ash (second), the nuclear energy generation rate
  (third), and the velocity out of the simulation plane (bottom).  The
  out of plane velocity contours indicate where the Coriolis force has
  created a hurricane-like wind structure which confines the flame,
  leading to an inclined front. It is at such an interface that the
  temperature and burning rate are highest.  }
\end{figure}

@@

\ack The work at Stony Brook was supported by DOE/Office of Nuclear
Physics grant DE-FG02-87ER40317 and contract 7418390 with Lawrence
Berkeley National Laboratory as part of the Exascale Compute Project
ExaStar collaboration.  The work at LBNL was supported by the DOE
Office of Advanced Scientific Computing Research under Contract No,
DE-AC02-05CH11231. YC is supported by the European Union Horizon 2020
research and innovation programme under the Marie Sklodowska-Curie
Global Fellowship grant agreement No 703916.  An award of computer
time was provided by the Innovative and Novel Computational Impact on
Theory and Experiment (INCITE) program. This research used resources
of the Oak Ridge Leadership Computing Facility at the Oak Ridge
National Laboratory, which is supported by the Office of Science of
the U.S. Department of Energy under Contract No.\ DE-AC05-00OR22725.
This research used resources of the National Energy Research
Scientific Computing Center, which is supported by the Office of
Science of the U.S. Department of Energy under Contract
No.\ DE-AC02-05CH11231.  Visualizations were done using yt~\cite{yt}.
This research has made use of NASA's Astrophysics Data System
Bibliographic Services.
